
---

# Background: Gaussian LDA

### Model

Assume the samples are distributed in $K$ groups and note

- $\mathbf{G}$ the group membership matrix
- $\mathbf{U} = [{\boldsymbol\mu}_1, \dots, {\boldsymbol\mu}_K]$ the matrix of group-specific means

The model is

$$\mathbf{Z}_i = \mathcal{N}(\mathbf{G}_i^\top \mathbf{U}, {\boldsymbol\Sigma})$$

$\rightsquigarrow$ Aim of LDA: Find the linear combination(s) $\mathbf{Z} u$ maximizing separation between groups


### Solution

Find the first eigenvectors of $\mathbf{\Sigma}_w^{-1} \mathbf{\Sigma}_b$ where

- $\mathbf{\Sigma}_w$ is the _within_-group variance matrix, i.e. the unbiased estimated of ${\boldsymbol\Sigma}$:
- $\mathbf{\Sigma}_b$ is the _between_-group variance matrix, estimated from $\mathbf{U}$.

---

# Poisson lognormal LDA (1)

Similar to normal PLN with

- $\mathbf{X} \rightarrow (\mathbf{X}, \mathbf{G})$
- $\mathbf{B} \rightarrow (\mathbf{B}, \mathbf{U})$

### Inference

-  Use .important[variational inference] to estimate $(\mathbf{B}, \mathbf{U})$, $\mathbf{\Sigma}$ and $\mathbf{Z}_i$

- Compute $\hat{\mathbf{\Sigma}}_b$ as

$$\hat{\mathbf{\Sigma}}_b = \frac1{K-1} \sum_k n_k (\hat{{\boldsymbol\mu}}_k - \hat{{\boldsymbol\mu}}_\bullet) (\hat{{\boldsymbol\mu}}_k - \hat{{\boldsymbol\mu}}_\bullet)^\intercal$$

- Compute first $K-1$ eigenvectors of $\hat{{\boldsymbol\Sigma}}^{-1} \hat{\mathbf{\Sigma}}_b = \mathbf{P}^\top \Lambda \mathbf{P}$


---

# Poisson lognormal LDA (2)

### Graphical representation

Mimick Gaussian LDA:

- Center the estimated latent positions $\tilde{\mathbf{Z}}$
- Compute the estimated coordinates along the discriminant axes

$$\tilde{\mathbf{Z}}^{LDA} = \tilde{\mathbf{Z}} \mathbf{P} \Lambda^{1/2}$$

### Prediction

For each group $k$
- Assume that the new sample $\mathbf{Y}_{\text{new}}$ comes from group $k$
- Compute (variational) _likelihood_ $p_k = \mathbb{P}(\mathbf{Y}_{\text{new}} | \hat{{\boldsymbol\Sigma}}, \hat{\mathbf{\Sigma}_b}, \hat{{\boldsymbol\mu}}_k)$
- Compute posterior probability $\pi_k \propto \frac{n_k p_k}{n}$

$\rightsquigarrow$ Assign to group with highest $\pi_k$


---

# Discriminant Analysis  (scRNA, 1)

Use the `cell-line` variable for grouping (`grouping` is a factor of group to be considered)

```{r PLNLDA-oaks, cache=TRUE, results=FALSE}
myLDA_cell_line <- 
  PLNLDA(counts ~ 1 + offset(log(total_counts)), grouping = cell_line,
         data = scRNA_train)
```

.pull-left[
```{r plot-lda-oaks1, echo = FALSE, fig.height = 6}
plot(myLDA_cell_line, map = "individual")
```
]

.pull-right[
```{r plot-lda-oaks2, echo = FALSE, fig.height = 6}
plot(myLDA_cell_line, map = "variable")
```
]

---
# Discriminant Analysis  (scRNA, 2)

Consider now a diagonal covariance.

```{r PLNLDA-oaks-diag, cache=TRUE, results=FALSE}
myLDA_cell_line_diag <- 
  PLNLDA(counts ~ 1 + offset(log(total_counts)), grouping = cell_line,
         data = scRNA_train, control = PLN_param(covariance = "diagonal"))
```

.pull-left[
```{r plot-lda-diag-oaks1, echo = FALSE, fig.height = 6}
plot(myLDA_cell_line_diag, map = "individual")
```
]

.pull-right[
```{r plot-lda-diag-oaks2, echo = FALSE, fig.height = 6}
plot(myLDA_cell_line_diag, map = "variable")
```
]

---
# Discriminant Analysis (scRNA, 3)

We can prediction cell-line of some new data: let us try either diagonal or fully parametrized covariance.

```{r, pred-scRNA, cache = TRUE}
pred_cell_line      <- predict(myLDA_cell_line     ,
                               newdata = scRNA_test, type = "response")
pred_cell_line_diag <- predict(myLDA_cell_line_diag,
                               newdata = scRNA_test, type = "response")
```

The ARI on the test set is impressive.footnote[The problem should be easy though...]

```{r, ari-lda-pred-scRNA}
aricode::ARI(pred_cell_line, scRNA_test$cell_line)
aricode::ARI(pred_cell_line_diag, scRNA_test$cell_line)
```

---
# Discriminant Analysis  (scRNA, 4)

Let us explore the discriminant groups: 

```{r LDA-heatmap-factors, fig.align='center'}
heatmap(exp(myLDA_cell_line$group_means))
```

Indeed, some groups of gene caracterize well the cell-lines.
