
---

# Poisson Lognormal PCA

### Model
  
$$\begin{array}{rcl}
  \mathbf{Z}_i & \sim^\text{iid} \mathcal{N}_p({\boldsymbol 0}_p, {\boldsymbol\Sigma}), & \text{rank}({\boldsymbol\Sigma}) = k \ll p \\
  \mathbf{Y}_i \,|\, \mathbf{Z}_i & \sim \mathcal{P}(\exp\{\mathbf{O}_i + \mathbf{X}_i \mathbf{B} + \mathbf{Z}_i\})
\end{array}$$

Recall that: $\text{rank}({\boldsymbol\Sigma}) = q \quad \Leftrightarrow \quad \exists \mathbf{C} (p \times q): \Sigma = \mathbf{C} \mathbf{C}^\intercal$.
  
### Estimation

Variational inference

$$\text{maximize } J({\boldsymbol\beta}, {\boldsymbol\psi}) $$
$\rightsquigarrow$ Still bi-concave in ${\boldsymbol\beta} = (\mathbf{C}, \mathbf{B})$ and ${\boldsymbol\psi} = (\mathbf{M}, \mathbf{S})$


---

# Model selection and Visualization

### Number of components/rank $k$ needs to be chosen.
  
$\log p_{\hat{\boldsymbol\beta}}(\mathbf{Y})$ intractable: use variational "likelihood"  $J(\hat{\boldsymbol\beta}, \hat{\boldsymbol\psi})$

- BIC $\rightsquigarrow$ $\text{vBIC}_k = J(\hat{\boldsymbol\beta}, \tilde{p}) - \frac12 p (d + k) \log(n)$
- ICL $\rightsquigarrow$ $\text{vICL}_k = \text{vBIC}_k - \mathcal{H}(\tilde{p})$

$$
  \hat{k} = \arg\max_k \text{vBIC}_k
  \qquad \text{or} \qquad
  \hat{k} = \arg\max_k \text{vICL}_k
$$
### Visualization

- Gaussian PCA: Optimal subspaces nested when $q$ increases.
- PLN-pPCA: Non-nested subspaces.

For the selected dimension dimension $\hat{k}$:

- Compute the estimated latent positions $\mathbb{E}_q(\mathbf{Z}_i) = \mathbf{M} \hat{\mathbf{C}}^\top$
- Perform PCA on the $\mathbf{M} \hat{\mathbf{C}}^\top$
- Display result in any dimension $k \leq \hat{k}$

---

# PCA: Goodness of fit

.important[pPCA:] Cumulated sum of the  eigenvalues = \% of variance preserved on the first $q$ components.

### PLN-pPCA

Deviance based criterion.

- Compute $\tilde{\mathbf{Z}}^{(k)} = \mathbf{O} + \mathbf{X} \hat{\mathbf{B}}^\top + \mathbf{M}^{(k)} \left(\hat{\mathbf{C}}^{(k)}\right)^\top$
- Take $\lambda_{ij}^{(k)} = \exp\left(\tilde{Z}_{ij}^{(k)}\right)$
- Define $\lambda_{ij}^{\min} = \exp( \tilde{Z}_{ij}^0)$ and $\lambda_{ij}^{\max} = Y_{ij}$
- Compute the Poisson log-likelihood $\ell_k = \log \mathbb{P}(\mathbf{Y}; \lambda^{(k)})$

### Pseudo R²

$$R_k^2 = \frac{\ell_k - \ell_{\min}}{\ell_{\max} - \ell_{\min}}$$

---
# A PCA analysis of the scRNA data set (1)

```{r PLNPCA offset, cache=TRUE, warning=FALSE, message=FALSE, results='hide'}
PCA_scRNA <- PLNPCA(counts ~ 1 + offset(log(total_counts)), data = scRNA_train,
                    ranks = c(1, 2, seq(5, 40, 5)))
```

### Model Selection
```{r PCA offset vizu cell-line, fig.width=6, fig.height=6, fig.align="center", echo=FALSE}
plot(PCA_scRNA)
myPCA_scRNA <- getBestModel(PCA_scRNA, "BIC")
```

---
# A PCA analysis of the scRNA data set  (2)

### Biplot

Individual + Factor map - 40 most contributing genes

<small>
```{r PCA offset vizu tree, fig.width=6, fig.height=6, fig.align="center", echo=FALSE}
myPCA_scRNA %>% factoextra::fviz_pca_biplot(
  select.var = list(contrib = 40), addEllipses = TRUE, habillage = scRNA_train$cell_line,
  title = "Biplot (50 most contributing genes)"
  ) + labs(col = "tree status") + scale_color_viridis_d()
```
</small>

