---
title: "Poisson lognormal models for count data"
subtitle: "Focus on some probabilistic factor models: Poisson LDA and Poisson PCA"
author: "J. Chiquet, M. Mariadassou, S. Robin, B. Batardière + others<br /><br /> <small>MIA Paris-Saclay, AgroParisTech, INRAE, Sorbonne Universyty </small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://pln-team.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: inverse, middle

# Outline 

  1. .large[**Framework** of multivariate Poisson lognormal models] 
  <br/>

  2. .large[**PLN estimation** with variational inference] 
  <br/>

  3. .large[**Focus** on factor analysis with PLN (LDA, PCA)] 
  <br/>
  
  4. .large[**Illustration** on a scRNA dataset] 

---
class: inverse, center, middle

# Multivariate Poisson lognormal models <br/> .small[Motivations, Framework]

```{r setup, echo=FALSE, cache=FALSE, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR) # reference manager
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)

library(tidyverse)  # data manipulation
library(PLNmodels)  # plot of covariance/correlation matrices
library(corrplot)   # plot of covariance/correlation matrices

options(future.fork.enable = TRUE)
future::plan("multicore", workers = 10)
```

<!-- STATISTICAL MODEL -->
```{r PLN_stat_model, child='PLN_stat_model.Rmd'}
```

---
class: inverse, center, middle

# Variational inference for standard PLN<br/> .small[Optimisation]

<!-- VARIATIONAL INFERENCE -->
```{r PLN_var_inference, child='PLN_var_inference.Rmd'}
```

```{r PLN_scRNA_analysis, child='PLN_scRNA_analysis.Rmd'}
```


---
class: inverse, center, middle

# Probabilistic factor models in PLN<br/> .small[Poisson LDA and Poisson PCA]

---

# Natural extensions of PLN

### Various tasks of multivariate analysis

  - .important[Dimension Reduction]: rank constraint matrix $\boldsymbol\Sigma$.
    
    $$\mathbf{Z}_i \sim \mathcal{N}(\boldsymbol\mu, \boldsymbol\Sigma = \mathbf{C}\mathbf{C}^\top), \quad \mathbf{C} \in \mathcal{M}_{pk} \text{ with orthogonal columns}.$$

  - .important[Classification]: maximize separation between groups with means 

  $$\mathbf{Z}_i \sim \mathcal{N}(\sum_k {\boldsymbol\mu}_k \mathbf{1}_{\{i\in k\}}, \boldsymbol\Sigma), \quad \text{for known memberships}.$$

  - .important[Clustering]: mixture model in the latent space 

  $$\mathbf{Z}_i \mid i \in k  \sim \mathcal{N}(\boldsymbol\mu_k, \boldsymbol\Sigma_k), \quad \text{for unknown memberships}.$$

  - .important[Network inference]: sparsity constraint on inverse covariance.
  
  $$\mathbf{Z}_i \sim \mathcal{N}(\boldsymbol\mu, \boldsymbol\Sigma = \boldsymbol\Omega^{-1}), \quad \|\boldsymbol\Omega \|_1 < c.$$

  - .important[Variable selection]: sparsity constraint on regression coefficients
  
  $$\mathbf{Z}_i \sim \mathcal{N}(\mathbf{x}_i^\top\mathbf{B}, \boldsymbol\Sigma), \quad \|\mathbf{B} \|_1 < c.$$


---

# Natural extensions of PLN

### Tasks seen in Factor Analysis

  - .important[Dimension Reduction]: rank constraint matrix $\boldsymbol\Sigma$. .important[PCA]
    
    $$\mathbf{Z}_i \sim \mathcal{N}(\boldsymbol\mu, \boldsymbol\Sigma = \mathbf{C}\mathbf{C}^\top), \quad \mathbf{C} \in \mathcal{M}_{pk} \text{ with orthogonal columns}.$$

  - .important[Classification]: maximize separation between groups with means .important[LDA]

  $$\mathbf{Z}_i \sim \mathcal{N}(\sum_k {\boldsymbol\mu}_k \mathbf{1}_{\{i\in k\}}, \boldsymbol\Sigma), \quad \text{for known memberships}.$$


```{r PLN_LDA, child='PLN_LDA.Rmd'}
```

```{r PLN_PCA, child='PLN_PCA.Rmd'}
```

---
class: inverse, center, middle

# On-going works

---
# Optimisation, Algorithms

With .important[Bastien Batardière, Joon Kwon, Julien Stoehr]

## Exact Maximization

Direct likelihood optim (SGD with Important Sampling)

## Optimization

Optimisation guarantees coupling adaptive SGD + variance reduction

---
# PLN LDA: Ongoing Extension

With .important[Nicolas Jouvin]

### Quadratic Discriminant Analysis

- Relax assumption of common covariance between groups
- Tests Linear vs Quadratic Discriminant

### Regularized Discriminant Analysis

- Ridge-like regularized Covariance
- Sparse Covariance
- Block-wise Covariance

$\rightsquigarrow$ better assess the structure of the sub-population

---
# PLN PCA: Ongoing Extension

With .important[Nicolas Jouvin]

## Mixture of PLN PCA

- Assume several sub-population in the latent space
- Each sub-population has is represented by a different linear combinasion of the initial variables
- In the PLN setup, several variational approximation are possible

## Functional PCA

PhD of Barbara Bricout (S. Robin, S. Donnet)

Features are time points

---
# Temporal/Spatial dependencies

With .important[Stéphane Robin, Mahendra Mariadassou]

## 'In-row' dependency structure

Consider, e.g., a multivariate Gaussian auto-regressive process on the latent variable $\mathbf{Z}$

$$\mathbf{Z}_1 \sim \mathcal{N}(0, {\boldsymbol\Sigma}), \quad \mathbf{Z}_t = \mathbf{A} \mathbf{Z}_{t-1} + {\boldsymbol\varepsilon}_t \quad
({\boldsymbol\varepsilon})_t \text{ iid } \sim \mathcal{N}(0, \Delta^{-1})$$

The counts are described by $(\mathbf{Y}_t) \in \mathbb{N}^p$ with time $t\geq 0$

$$(\mathbf{Y}_t) \text{ indep. } \mathbf{Z}_t: \quad \mathbf{Y}_t \sim \mathcal{P}\left(\exp\left\{ \mathbf{x}_t^\top \mathbf{B} + \mathbf{Z}_t\right\}\right)$$

$\rightsquigarrow$ Camille Mondon's MsC

## Inference

Based on Kalman Filter in the latent layer + variational approximation

---

# A zero-inflated PLN

With .important[Bastien Batardière, Mahendra Mariadassou]

### Motivations 

- account for a large amount of zero, i.e. with single-cell data,
- try to separate "true" zeros from "technical"/dropouts

### The Model

Use two latent vectors $\mathbf{W}_i$ and $\mathbf{Z}_i$ to model excess of zeroes and dependence structure 
$$\begin{aligned}
   \mathbf{Z}_i & \sim \mathcal{N}({\mathbf{o}_i + \mathbf{x}_i^\top\mathbf{B}},\boldsymbol\Sigma) \\
    W_{ij} & \sim \mathcal{B}(\text{logit}^{-1}(\mathbf{x}_i^\top{\mathbf{B}}_j^0)) \\
    Y_{ij} \, | \, W_{ij}, Z_{ij} & \sim W_{ij}\delta_0 + (1-W_{ij}) \mathcal{P}\left(\exp\{Z_{ij}\}\right), \\
\end{aligned}$$
  
The unkwown parameters are 
- $\mathbf{B}$, the regression parameters (from the PLN component)
- $\mathbf{B}^0$, the regression parameters (from the Bernoulli component)
- $\boldsymbol\Sigma$, the variance-covariance matrix

$\rightsquigarrow$ ZI-PLN is a mixture of PLN and Bernoulli distribution with shared covariates.

---

# ZI-PLN Inference

### Variational approximation 1.

\begin{equation*}
p(\mathbf{Z}_i, \mathbf{W}_i  \mathbf{Y}_i) \approx q_{\psi}(\mathbf{Z}_i, \mathbf{W}_i) \approx q_{\psi_1}(\mathbf{Z}_i) q_{\psi_2}(\mathbf{W}_i)
\end{equation*}

with

\begin{equation*}
q_{\psi_1}(\mathbf{Z}_i) = \mathcal{N}(\mathbf{Z}_i; \mathbf{m}_{i}, \mathrm{diag}(\mathbf{s}_{i} \circ \mathbf{s}_{i})), \qquad q_{\psi_2}(\mathbf{W}_i) = \otimes_{j=1}^p \mathcal{B}(W_{ij}, \pi_{ij})
\end{equation*}

.important[Tested, works _partially_] Too rough variational approximation

### Variational approximation 1.

\begin{equation*}
p(\mathbf{Z}_i, \mathbf{W}_i  \mathbf{Y}_i) \approx q_{\psi}(\mathbf{Z}_i, \mathbf{W}_i) \approx q_{\psi_1}(\mathbf{Z}_i|\mathbf{W}_i) q_{\psi_2}(\mathbf{W}_i)
\end{equation*}

\begin{equation*}
q_{\psi_1}(\mathbf{Z}_i | \mathbf{W}_i) = \text{ mixture of Gaussians}, \quad  q_{\psi_2}(\mathbf{W}_i) = \otimes_{j=1}^p \mathcal{B}(W_{ij}, \pi_{ij})
\end{equation*}


---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Flexible modeling of the covariance structure, allows for covariates
  - Efficient V-EM algorithm

## Advertisement 

[https://computo.sfds.asso.fr](https://computo.sfds.asso.fr), a journal promoting reproducible research in ML and stat.

---
# References

```{r, results='hide', echo = FALSE}
RefManageR::Cite(myBib, 'PLNmodels')
RefManageR::Cite(myBib, 'PLNPCA')
RefManageR::Cite(myBib, 'PLNnetwork')
RefManageR::Cite(myBib, 'Facon2021')
RefManageR::Cite(myBib, 'Lejal2021')
```

<small>
```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```
</small>

