---
title: "Poisson lognormal models for count data"
subtitle: "Focus on "
author: "J. Chiquet, M. Mariadassou, S. Robin, B. Batardière + others<br /><br /> <small>MIA Paris-Saclay, AgroParisTech, INRAE, Sorbonne Universyty </small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://pln-team.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: inverse, middle

# Outline 

  1. .large[**Framework** of multivariate Poisson lognormal models] 
  <br/>

  2. .large[**Focus** on factor analysis with PLN (PCA, LDA)] 
  <br/>
  
  3. .large[**Optimization** with Variational inference] 
  <br/>
  
  4. .large[**Ongoing extension** for factor models] 

---
class: inverse, center, middle

# Multivariate Poisson lognormal models <br/> .small[Motivations, Framework]

```{r setup, echo=FALSE, cache=FALSE, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(RefManageR) # reference manager
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)
```

<!-- STATISTICAL MODEL -->
```{r PLN_stat_model, child='PLN_stat_model.Rmd'}
```

<!-- ILLUSTRATION -->
```{r PLN_scRNA_analysis, child='PLN_scRNA_analysis.Rmd'}
```

---
class: inverse, center, middle

# Variational inference for standard PLN<br/> .small[Optimisation]

<!-- VARIATIONAL INFERENCE -->
```{r PLN_var_inference, child='PLN_var_inference.Rmd'}
```

---
class: inverse, center, middle

# Some on-going extensions


---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Flexible modeling of the covariance structure, allows for covariates
  - Efficient V-EM algorithm
  - Variational estimator is asymptotically normal (and hopefully unbiased) with computable covariance matrix. 
  - ZI-PLN reduces (some) problems induced by high sparsity in the data

## Work in progress

  - Caracterisation of Variational Estimator
  - with .important[J. Stoehr] Direct likelihood optim (SGD with Important Sampling)
  - with .important[J. Kwon]: optimisation guarantees coupling adaptive SGD + variance reduction
  - Connection/Comparison with VAE with e.g Poisson neg log-likelihood as loss
  
## Advertisement 

[https://computo.sfds.asso.fr](https://computo.sfds.asso.fr), a journal promoting reproducible research in ML and stat.

---
# References

```{r, results='hide', echo = FALSE}
RefManageR::Cite(myBib, 'PLNmodels')
RefManageR::Cite(myBib, 'PLNPCA')
RefManageR::Cite(myBib, 'PLNnetwork')
RefManageR::Cite(myBib, 'Facon2021')
RefManageR::Cite(myBib, 'Lejal2021')
```

<small>
```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```
</small>

