
---
# Poisson Lognormal model for PCA

The PLN-PCA `r Citep(myBib, "PLNPCA")` model implemented in *PLNmodels* can viewed as a PLN model with an additional rank constraint on the covariance matrix $\boldsymbol\Sigma$ such that $\mathrm{rank}(\boldsymbol\Sigma)= q$:

$$\begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}(\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta,\boldsymbol\Sigma), & \boldsymbol\Sigma = \mathbf{B}\mathbf{B}^\top, \quad \mathbf{B}\in\mathcal{M}_{pq} \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} & Y_{ij} | Z_{ij} \sim \mathcal{P}\left(\exp\{Z_{ij}\}\right),
\end{array}$$

The dimension $q$ of the latent space corresponds to the number of axes in the PCA or, in other words, to the rank of $\boldsymbol\Sigma = \mathbf{B}\mathbf{B}^\intercal$.


The unkwown parameters are  $\boldsymbol\Theta$ and $\mathbf{B}$, the matrix of .important[_rescaled loadings_]

### Features

  - **Optimization**: _"Similar"_ variational framework (with different gradients)
  - **Model selection**: variational BIC/ICL
      - $\tilde{\text{BIC}}_q = J(\theta, q) - \frac12 \log(n) \left(p (d + q) - q(q-1)/2\right)$
      - $\tilde{\text{ICL}}_q = \tilde{\text{BIC}}_q - \mathcal{H}(q)$
  - **Vizualization:** PCA on the expected latent position $\mathbb{E}_{q}(\mathbf{Z}_i) -\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta  = \mathbf{M}\hat{\mathbf{B}}^\top$

---
# A PCA analysis of the oaks data set

Let us fit PLNPCA on our best model up to now (with TSS as offsets):

```{r PLNPCA offset, cache = TRUE, results = FALSE}
PCA_offset <- 
  PLNPCA(Abundance ~ 1 + offset(log(Offset)), data = oaks, ranks = 1:30)
```


The ouput is of class (`PLNPCAfamily`). It a collection of `R` objects:

```{r print PLNPCA offset}
PCA_offset
```

`PLNPCAfamily` has three methods: `plot`, `getModel`, `getBestModel`<sup>1</sup> 

.footnote[[1] Additional help can be found with `?PLNPCAfamily`, `?getBestModel.PLNPCAfamily`, `?plot.PLNPCAfamily`]

---
# PCA analysis: model selection (I)

The plot function gives you hints about the "right" rank/subspace size of your data

```{r plot PLNPCA offset, fig.width = 12, fig.height=6}
plot(PCA_offset, reverse = TRUE)
```

---
# PCA analysis: model selection (II)

To extract a particular model from the collection, use `getBestModel`:

```{r extract PLNPCA offset}
PCA_offset_BIC <- getBestModel(PCA_offset, "BIC")
```


The extracted object has class `PLNPCAfit`. It inherits from the `PLNfit` class but with additional methods due to its `PCA` nature: when printing `PCA_offset_BIC`, we get

```{r print PLNPCAfit, echo = FALSE}
PCA_offset_BIC
```

---
# PCA analysis: model exploration

Inheritance allows you to rely on the same methods as with `PLN`:

.pull-left[
```{r PLN covariance M01}
corrplot(
  cov2cor(sigma(M01_oaks)),
  tl.cex = .5)
```
]

.pull-right[
```{r PLNPCA covariance}
corrplot(
  cov2cor(sigma(PCA_offset_BIC)),
  tl.cex = .5)
```
]

---
# PCA: vizualisation 

<small>
```{r PCA offset vizu tree, fig.width=6, fig.height=6, fig.align="center"}
factoextra::fviz_pca_biplot(
  PCA_offset_BIC, select.var = list(contrib = 10), addEllipses = TRUE, habillage = oaks$tree,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "tree status") + scale_color_viridis_d()
```
</small>

---
# PCA: removing covariate effects

To hopefully find some hidden effects in the data, we can try to remove confounding ones:

```{r PCA covariate tree, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
PCA_tree <- PLNPCA(Abundance ~ 0 + tree + offset(log(Offset)), 
                   data = oaks, ranks = 1:30)
```

```{r PCA covariate tree plot, echo = FALSE, fig.align="center", fig.width=6, fig.height=6}
PCA_tree %>% getBestModel("BIC") %>% 
factoextra::fviz_pca_biplot(
  select.var = list(contrib = 10), col.ind = oaks$distTOground,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "distance (cm)") + scale_color_viridis_c()
```
