
---

# Sparse precision for multivariate counts

The PLN-network model add a sparsity constraint on the precision matrix ${\boldsymbol\Sigma}^{-1}\triangleq \boldsymbol\Omega$:

$$\begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}\left({\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta},\boldsymbol\Omega^{-1}\right) &  \|\boldsymbol\Omega\|_1 < c \\
    \text{observation space } &  \mathbf{Y}_i | \mathbf{Z}_i \sim \mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right)
  \end{array}$$


$\rightsquigarrow$ The $\ell_1$-penalty induces selection of direct relations (an underlying network)

## .small[Variational approximation]
  
$$J(\theta, q)  - \lambda  \| \boldsymbol\Omega\|_{1,\text{off}} = \mathbb{E}_{q} [\log p_\theta(\mathbf{Y}, \mathbf{Z})] + \mathcal{H}[q(\mathbf{Z})] - \lambda  \|\boldsymbol\Omega\|_{1, \text{off}}$$

Still bi-concave in $(\boldsymbol\Omega, \boldsymbol\Theta)$ and $(\mathbf{M}, \mathbf{S})$.  

Solving in  $\boldsymbol\Omega$ leads to

$$\hat{\boldsymbol\Omega} = \arg\max_{\boldsymbol\Omega} \frac{n}{2} \left(\log | \boldsymbol\Omega | - \text{trace}(\hat{\boldsymbol\Sigma} \boldsymbol\Omega)\right) - \lambda \|\boldsymbol\Omega\|_{1, \text{off}}: \quad \text{graphical-Lasso problem}$$
with $\hat{\boldsymbol\Sigma} = n^{-1}(\mathbf{M}^\top \mathbf{M} + \mathrm{diag}(\bar{\mathbf{S}}^2)$.

---
# Network inference on the oaks data set

We try the inference of a collections of networks, accounting for the tree effect.

Models are indexed by the sparsity level (or penalty). (E)BIC or similar information criteria can help choosing the right amount.

.pull-left[
```{r PLNnetwork tree, cache = TRUE, results = FALSE}
networks_oaks_tree <- 
  PLNnetwork(
    Abundance ~ 0 + tree + 
           offset(log(Offset)),
    data = oaks
  )
```
]

.pull-right[
```{r plot network family site, fig.width = 7, fig.height = 7, echo = FALSE, caption = "with tree correction"}
plot(networks_oaks_tree)
```
]

---
# PLNnetwork: field access

Let us plot the estimated correlation matrix, after regularization of its inverse, and the corresponding network of partial correlation.

.pull-left[
```{r plot network site, fig.width = 7, fig.height = 7, caption = "with site correction", echo = FALSE}
net2 <- getBestModel(networks_oaks_tree)
corrplot(cov2cor(sigma(net2)), cl.pos = "n",
  is.corr = FALSE, method = 'color')
```
]

.pull-right[
```{r plot net site, fig.width = 8, fig.height = 8, caption = "with site correction", echo = FALSE}
plot(net2)
```
]

---
# PLNnetwork: stability selection

An alternative to model selection criteria is the stability selection  - or StARS in the context of network.

- Basically, it uses resampling to estimate robustness of each edges in the network.
- we keep the value of the penalty that guarantees a given level of robustness in the whole network

.important[Careful!]: its is computationally expensive.

--

In `getBestModel`, when "StARS" is requested, stabiltiy selection is performed if needed:

```{r stability selection, cache = TRUE, eval = FALSE}
net3 <- getBestModel(networks_oaks_tree, "StARS") 
```
