
---

# The PLN function

The `PLN` function works in the same way as `lm`: 

```{r, eval = FALSE}
PLN(formula = , # mandatory
    data    = , # highly recommended
    subset    , # optional  
    weights   , # optional 
    control     # optional, mostly for advanced users
    )
```

- `data` specifies where to look for the variables
- `formula` specifies the relationship between variables in `data`

  ( $\rightsquigarrow$ _It builds matrices_ $\mathbf{Y},\mathbf{X},\mathbf{O}$)

- `subset` is used for subsampling the observations, it should be a .important[full length] boolean vector, not a vector of indices / sample names
- `weights` is used to weighting the observations, 
- `control` is (mainly) used for tuning the optimization and should typically not be changed.
---

# Simple PLN models on oaks data

The simplest model we can imagine only has an intercept term:

```{r oaks PLN, cache = TRUE, results = FALSE}
M00_oaks <- PLN(Abundance ~ 1, oaks)
```

`M00_oaks` is a particular `R` object with class `PLNfit` that comes with a couple methods, helpfully listed when you print the object :

```{r oaks PLN print}
M00_oaks
```

---

# Accessing parameters

```{r simple PLN coeff}
coef(M00_oaks) %>% head() %>% t() %>% knitr::kable(format = "html")
```

.pull-left[
```{r simple PLN covariance, fig.height=5}
sigma(M00_oaks) %>% 
  corrplot(is.corr=FALSE, tl.cex = .5)
```
]

.pull-right[
```{r simple PLN correlation, fig.height=5}
sigma(M00_oaks) %>% cov2cor() %>% 
  corrplot(tl.cex = .5)
```
]

---
#  Adding Offsets and covariates

## Offset: .small[modeling sampling effort]

The predefined offset uses the total sum of reads, accounting for technologies specific to fungi and bacteria:

```{r simple PLN offsets, cache = TRUE, results = FALSE}
M01_oaks <- PLN(Abundance ~ 1 + offset(log(Offset)) , oaks)
```

## Covariates: .small[tree and orientation effects ('ANOVA'-like) ]

The `tree` status is a natural candidate for explaining a part of the variance.

- We chose to describe the tree effect in the regression coefficient (mean)
- A possibly spurious effect regarding the interactions  between species (covariance).

```{r PLN covariate oaks, cache = TRUE, results = FALSE}
M11_oaks <- PLN(Abundance ~ 0 + tree + offset(log(Offset)), oaks)
```

What about adding more covariates in the model, e.g. the orientation?

```{r PLN regroup oaks modalities, cache = TRUE, results = FALSE}
M21_oaks <- PLN(Abundance ~  0 + tree + orientation + offset(log(Offset)), oaks)
```

---
#  Adding Offsets and covariates (II)

There is a clear gain in introducing the tree covariate in the model:

```{r PLN covariate oaks results}
rbind(M00 = M00_oaks$criteria, M01 = M01_oaks$criteria,
      M11 = M11_oaks$criteria, M21 = M21_oaks$criteria) %>% 
  knitr::kable(format = "html")
```

Looking at the coefficients $\mathbf{\Theta}$ associated with `tree` bring additional insights:

```{r oaks matrix plot, fig.width=14, fig.height=2, echo = FALSE}
coef(M11_oaks) %>% t() %>% corrplot(method = "color", is.corr = FALSE, tl.cex = 1, cl.pos = "n")
```

---

# Residuals correlations

For models with offsets and tree effect in the mean:

.pull-left[
```{r simple PLN covariance M01, fig.height=5}
sigma(M01_oaks) %>% cov2cor() %>% 
  corrplot(is.corr=FALSE, tl.cex = .5)
```
]

.pull-right[
```{r simple PLN correlation M11, fig.height=5}
sigma(M11_oaks) %>% cov2cor() %>% 
  corrplot(tl.cex = .5)
```
]
