---
title: "PLNmodels"
subtitle: "A collection of Poisson lognormal models <br/> for multivariate analysis of count data"
author: "J. Chiquet, M. Mariadassou, S. Robin<br /><br /> <small>INRA - Applied Mathematics and Informatics Division</small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://jchiquet.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse) # data manipulation
library(corrplot)  # plot of covariance/correlation matrices
```

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)
```

class: inverse, center, middle

# Getting Started

---
# Requirements

## Package PLNmodels

You can install the last stable release of **PLNmodels** from the CRAN. The development version is available from GitHub.

```{r install, eval=FALSE, tidy=FALSE}
install.packages("PLNmodels")
devtools::install_github("jchiquet/PLNmodels")
```

## Dependencies

Make sure the following CRAN packages are installed:

```{r packages dependencies, eval=FALSE, tidy=FALSE}
required_CRAN <- c("R6", "glassoFast", "Matrix", "Rcpp", "RcppArmadillo",
                   "nloptr", "igraph", "grid", "gridExtra", "dplyr",
                   "tidyr", "purrr", "ggplot2", "corrplot", "magrittr", "rlang")

not_installed_CRAN <- setdiff(required_CRAN, rownames(installed.packages()))
if (length(not_installed_CRAN) > 0) install.packages(not_installed_CRAN)
```

---

# First steps

## Loading the package

Check that the installation process succeeded

```{r loading package}
library(PLNmodels)
packageVersion("PLNmodels")
```

## Finding help and documentation

The [PLNmodels website](https://jchiquet.github.io/PLNmodels/) contains

- the standard package documentation 
- a set of comprehensive vignettes for the top-level functions

all formatted with [**pkgdown**](https://pkgdown.r-lib.org)


And do not forget `?`/`??` in the `R` console.

---

class: inverse, center, middle

# 'Oaks powdery mildew' <br/> A motivating companion data set <br/> .small[See `r Citet(myBib, "JAL16")`]


---

# Generic form of data sets
  
Routinely gathered in ecology/microbiology/genomics 

## Data tables

  - .important[Abundances]: read counts of species/transcripts $j$ in sample $i$
  - .important[Covariates]: value of environmental variable $k$ in sample $i$
  - .important[Offsets]: sampling effort for species/transcripts $j$ in sample $i$

## Need for multivariate analysis

  - exhibit .important[patterns of diversity] <br/>
      $\rightsquigarrow$ summarize the information  (PCA, clustering, $\dots$)
  - understand .important[between-species interactions] <br />
      $\rightsquigarrow$ 'network' inference (variable/covariance selection)
  - correct for technical and .important[confounding effects] <br/>
      $\rightsquigarrow$ account for covariables and sampling effort

$\rightsquigarrow$ need a generic framework to _model dependences between count variables_

---

# 'oaks' data set overview

The `oaks` variable is loaded by the function `data()` (try `?oaks` to get additional details).
```{r load oaks}
data("oaks")
```

It consists in a special data frames ready to play with, typical from ecological data sets

<small>
```{r oaks struct}
str(oaks, max.level = 1)
```
</small>
---

# Abundance table (I)

```{r glimpse Abundance}
oaks$Abundance %>% as_tibble() %>% 
  dplyr::select(1:10) %>% 
  head() %>% DT::datatable(fillContainer = FALSE)
```

---

# Abundance table (II)

```{r glance Abundances, fig.height=6}
log(1 + oaks$Abundance) %>% 
  corrplot::corrplot(is.corr = FALSE,
    addgrid.col = NA,  tl.cex = .5,  cl.pos = "n")
```

---

# Covariates and offsets

Characterize the samples and the sampling, most important being

- `tree`: Tree status with respect to the pathogen (susceptible, intermediate or resistant)
- `distTOground`: Distance of the sampled leaf to the base of the ground
- `readsTOTfun`: Total number of ITS1 reads for that leaf
- `readsTOTbac`: Total number of 16S reads for that leaf

```{r covariates summary}
summary(oaks$tree)
summary(oaks$distTOground)
```

$\rightsquigarrow$ `readsTOTfun` and `readsTOTbac` are candidate for modeling sampling effort as offsets


---

class: inverse, center, middle

# Multivariate Poisson lognormal models <br/> statistical framework

---

# Models for multivariate count data

## If we were in a Gaussian world...

The .important[general linear model] `r Citep(myBib, "MKB79")` would be appropriate! For each sample $i = 1,\dots,n$, 

$$\underbrace{\mathbf{Y}_i}_{\text{abundances}} =  \underbrace{\mathbf{x}_i^\top \boldsymbol\Theta}_{\text{covariates}} + \underbrace{\mathbf{o}_i}_{\text{sampling effort}} + \boldsymbol\varepsilon_i, \quad \boldsymbol\varepsilon_i \sim \mathcal{N}(\mathbf{0}_p, \underbrace{\boldsymbol\Sigma}_{\text{between-species dependencies}})$$

null covariance $\Leftrightarrow$ independence $\rightsquigarrow$ uncorrelated species/transcripts do not interact

$\rightsquigarrow$ This model gives birth to Principal Component Analysis,  Discriminant Analysis, Gaussian Graphical Models, Gaussian Mixture models and many others $\dots$


## With count data...

There is no generic model for multivariate counts

  - Data transformation (log, $\sqrt{}$) : quick and dirty
  - Non-Gaussian multivariate distributions `r Citep(myBib, "IYA16")`: do not scale to data dimension yet
  - .important[Latent variable models]: interaction occur in a latent (unobserved) layer

---

# The Poisson Lognormal model (PLN)

The PLN model `r Citep(myBib, "AiH89")` is a .important[multivariate generalized linear model], where 

- the counts $\mathbf{Y}_i$ are the response variables
- the main effect is due to a linear combination of the covariates $\mathbf{x}_i$
- a vector of offsets $\mathbf{o}_i$ can be specified for each sample.

$$
\mathbf{Y}_i | \mathbf{Z}_i \sim \mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right), \qquad \mathbf{Z}_i \sim \mathcal{N}({\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta},\boldsymbol\Sigma), \\
$$

.pull-left[The unkwown parameters are 
- $\boldsymbol\Theta$, the regression parameters
- $\boldsymbol\Sigma$, the variance-covariance matrix
]

.pull-right[
PLN has some expected properties
- over-dispersion
- covariance with arbitrary signs
]


When all individuals $i=1,\dots,n$ are together, the data matrices required are
  - the $n\times p$ matrix of counts  $\mathbf{Y}$
  - the $n\times d$ matrix of design  $\mathbf{X}$
  - the $n\times p$ matrix of offsets $\mathbf{O}$

---
# Geometrical view

```{r PLN geometry, echo=FALSE, message=FALSE, fig.width=12, fig.height=8}
library(gridExtra)
set.seed(20171110)
x <- rnorm(100)
y <- rnorm(100)
b <- data.frame(x = x + y, y = y) / 1
mu <- 0
##
data.perfect <- as.data.frame((b + matrix(rep(mu, each = length(x)), ncol = 2)))
p.latent <- ggplot(data.perfect, aes(x, y)) + geom_point() + ggtitle(expression(Latent~Space~(Z)))
.rpois <- function(lambda) {
  unlist(lapply(exp(lambda), function(x) {rpois(1, x)}))
}
observation <- as.data.frame(lapply(data.perfect, .rpois))
mapped.parameter <- as.data.frame(lapply(data.perfect, exp))
## segment between mapped and observed data
segment.data <- cbind(mapped.parameter, observation)
names(segment.data) <- c("x", "y", "xend", "yend")
## Mapped parameters
p.mapped <- ggplot(mapped.parameter, aes(x, y)) + geom_point(col = "red") + ggtitle(expression(Observation~Space~(exp(Z))))
## Observations only
obs <- group_by(observation, x, y)
obs <- dplyr::summarize(obs, count = n())
p.observation.only <- ggplot(obs, aes(x, y)) +
  geom_point(aes(size = count)) +
  ggtitle(Observation~Space~(Y)~+'noise') +
  theme(legend.position = c(.95, .95), legend.justification = c(1, 1),
        legend.background = element_rect(color = "transparent"),
        legend.box.background = element_blank())
## Observations and latent parameters
p.observation.mixed <- p.observation.only +
  geom_point(data = mapped.parameter, color = "red", alpha = 0.5) +
  geom_segment(data = segment.data, aes(xend = xend, yend = yend), color = "black", alpha = 0.2) +
  ggtitle(Observation~Space~(Y==P(exp(Z)))~+'noise')
  grid.arrange(p.latent + labs(x = "species 1", y = "species 2"),
               p.mapped  + labs(x = "species 1", y = "species 2"),
               p.observation.mixed + labs(x = "species 1", y = "species 2"),
               p.observation.only + labs(x = "species 1", y = "species 2"),
               ncol = 2)
```

---
# .small[Inference: latent model but intractable EM]
  
## Aim of the inference

  - estimate $\theta = (\boldsymbol\Theta, \boldsymbol\Sigma)$ 
  - predict the $\mathbf{Z}_i$

## Maximum likelihood

PLN is an incomplete data model: try EM
$$\log p_\theta(\mathbf{Y}) = \mathbb{E}_p [\log p_\theta(\mathbf{Y}, \mathbf{Z}) \,|\, \mathbf{Y}] + \mathcal{H}[p_\theta(\mathbf{Z}\,|\,\mathbf{Y})]$$ 

EM requires to evaluate (some moments of) $p_\theta(\mathbf{Z} \,|\,  \mathbf{Y})$, but there is no close form!

## Solutions

  - `r Citep(myBib, "AiH89")` resort on numerical integration; `r Citep(myBib, "Kar05")` Monte-Carlo integration
  - Several heuristics, not always well motivated, found in the literature...
  - .important[Variational approach] `r Citep(myBib, "WaJ08")`: use a proxy of $p_\theta(\mathbf{Z}\,|\,\mathbf{Y})$.

---
# Variational approach: principle

.important[See the outstanding Stéphane Robin's Lecture]

## Idea

  - Find a proxy of the conditional distribution $p(\mathbf{Z}\,|\,\mathbf{Y})$:

$$q(\mathbf{Z}) \approx p_\theta(\mathbf{Z} | \mathbf{Y}).$$
  - Choose a convenient class of distribution $\mathcal{Q}$ and minimize a divergence

$$q(\mathbf{Z})^\star  \arg\min_{q\in\mathcal{Q}} D\left(q(\mathbf{Z}), p(\mathbf{Z} | \mathbf{Y})\right).$$

## Popular choice

The Küllback-Leibler divergence .small[(error averaged wrt the approximated distribution)]

$$KL\left(q(\mathbf{Z}), p(\mathbf{Z} | \mathbf{Y})\right) = \mathbb{E}_q\left[\log \frac{q(z)}{p(z)}\right] = \int_{\mathcal{Z}} q(z) \log \frac{q(z)}{p(z)} \mathrm{d}z.$$

---
# Variational approach: PLN

## Class of distribution: diagonal multivariate Gaussian

$$\mathcal{Q} = \Big\{q: \quad q(\mathbf{Z}) = \prod_i q_i(\mathbf{Z}_i), \quad q_i(\mathbf{Z}_i) = \mathcal{N}(\mathbf{Z}_i; \mathbf{m}_i, \mathbf{s}_i \circ \mathbf{s}_i) \Big\}$$

Maximize the ELBO (Evidence Lower BOund):

$$J(\theta, q) = \log p_\theta(\mathbf{Y}) - KL[q_\theta (\mathbf{Z}) ||  p_\theta(\mathbf{Z} | \mathbf{Y})]  = \mathbb{E}_{q} [\log p_\theta(\mathbf{Y}, \mathbf{Z})] + \mathcal{H}[q(\mathbf{Z})]$$

## Variational EM

  - VE step: find the optimal $q$ (here, $\{(\mathbf{m}_i, \mathbf{s}_i)\}_{i=1,\dots,n} = \{\mathbf{M}, \mathbf{S}\}$): 
$$q^h = \arg \max J(\theta^h, q) = \arg\min_{q \in \mathcal{Q}} KL[q(\mathbf{Z}) \,||\, p_{\theta^h}(\mathbf{Z}\,|\,\mathbf{Y})]$$
  - M step: update $\hat{\theta}^h$
$$\theta^h = \arg\max J(\theta, q^h) = \arg\max_{\theta} \mathbb{E}_{q} [\log p_{\theta}(\mathbf{Y}, \mathbf{Z})]$$

---
# .small[Application to the optimization of PLN models]
  
## Property of PLN variational approximation

The ELBO $J(\theta, q)$ is bi-concave, i.e.
  - concave wrt $q = (\mathbf{M}, \mathbf{S})$ for given $\theta$ 
  - convace wrt $\theta = (\boldsymbol\Sigma, \boldsymbol\Theta)$ for given $q$ 
but .important[not jointly concave] in general.

## Optimization

Gradient ascent for the complete set of parameters<sup>1</sup> $(\mathbf{M}, \mathbf{S}, \boldsymbol\Sigma, \boldsymbol\Theta)$

  - **algorithm**: conservative convex separable approximations `r Citet(myBib, "Svan02")` <br/>
  - **implementation**: `NLopt` nonlinear-optimization library `r Citet(myBib, "nlopt")` <br/>
  - **initialization**: LM after log-trasnformation applied independently on each variables + concatenation of the regression coefficients + Pearson residuals

.footnote[[1] Alternating between variational and model parameters is useless here <br/>
          [2] Optimizing on $\mathbf{S}$ such as $\mathbf{S} \circ \mathbf{S} = \mathbf{S}^2$ is the variance avoids positive constraints ]

---

class: inverse, center, middle

# Multivariate Poisson Regression with PLN

---

# The PLN function

The `PLN` function works in the same way as `lm`: 

```{r, eval = FALSE}
PLN(formula = , # mandatory
    data    = , # highly recommended
    subset    , # optional  
    weights   , # optional 
    control     # optional, mostly for advanced users
    )
```

- `data` specifies where to look for the variables
- `formula` specifies the relationship between variables in `data`

  ( $\rightsquigarrow$ _It builds matrices_ $\mathbf{Y},\mathbf{X},\mathbf{O}$)

- `subset` is used for subsampling the observations, it should be a .important[full length] boolean vector, not a vector of indices / sample names
- `weights` is used to weighting the observations, 
- `control` is (mainly) used for tuning the optimization and should typically not be changed.
---

# Simple PLN models on oaks data

The simplest model we can imagine only has an intercept term:

```{r oaks PLN, cache = TRUE, results = FALSE}
M00_oaks <- PLN(Abundance ~ 1, oaks)
```

`M00_oaks` is a particular `R` object with class `PLNfit` that comes with a couple methods, helpfully listed when you print the object :

```{r oaks PLN print}
M00_oaks
```

---

# Accessing parameters

```{r simple PLN coeff}
coef(M00_oaks) %>% head() %>% t() %>% knitr::kable(format = "html")
```

.pull-left[
```{r simple PLN covariance, fig.height=5}
sigma(M00_oaks) %>% 
  corrplot(is.corr=FALSE, tl.cex = .5)
```
]

.pull-right[
```{r simple PLN correlation, fig.height=5}
sigma(M00_oaks) %>% cov2cor() %>% 
  corrplot(tl.cex = .5)
```
]

---
#  Adding Offsets and covariates

## Offset: .small[modeling sampling effort]

The predefined offset uses the total sum of reads, accounting for technologies specific to fungi and bacteria:

```{r simple PLN offsets, cache = TRUE, results = FALSE}
M01_oaks <- PLN(Abundance ~ 1 + offset(log(Offset)) , oaks)
```

## Covariates: .small[tree and orientation effects ('ANOVA'-like) ]

The `tree` status is a natural candidate for explaining a part of the variance.

- We chose to describe the tree effect in the regression coefficient (mean)
- A possibly spurious effect regarding the interactions  between species (covariance).

```{r PLN covariate oaks, cache = TRUE, results = FALSE}
M11_oaks <- PLN(Abundance ~ 0 + tree + offset(log(Offset)), oaks)
```

What about adding more covariates in the model, e.g. the orientation?

```{r PLN regroup oaks modalities, cache = TRUE, results = FALSE}
M21_oaks <- PLN(Abundance ~  0 + tree + orientation + offset(log(Offset)), oaks)
```

---
#  Adding Offsets and covariates (II)

There is a clear gain in introducing the tree covariate in the model:

```{r PLN covariate oaks results}
rbind(M00 = M00_oaks$criteria, M01 = M01_oaks$criteria,
      M11 = M11_oaks$criteria, M21 = M21_oaks$criteria) %>% 
  knitr::kable(format = "html")
```

Looking at the coefficients $\mathbf{\Theta}$ associated with `tree` bring additional insights:

```{r oaks matrix plot, fig.width=14, fig.height=2, echo = FALSE}
coef(M11_oaks) %>% t() %>% corrplot(method = "color", is.corr = FALSE, tl.cex = 1, cl.pos = "n")
```

---

class: inverse, center, middle

# Dimension reduction and vizualisation with PLNPCA <br/> .small[See `r Citet(myBib, "PLNPCA")`]

---
# Poisson Lognormal model for PCA

The PLN-PCA `r Citep(myBib, "PLNPCA")` model implemented in *PLNmodels* can viewed as a PLN model with an additional rank constraint on the covariance matrix $\boldsymbol\Sigma$ such that $\mathrm{rank}(\boldsymbol\Sigma)= q$:

$$\begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}(\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta,\boldsymbol\Sigma), & \boldsymbol\Sigma = \mathbf{B}\mathbf{B}^\top, \quad \mathbf{B}\in\mathcal{M}_{pq} \\
  \text{observation space } &  Y_{ij} | Z_{ij} \quad \text{indep.} & Y_{ij} | Z_{ij} \sim \mathcal{P}\left(\exp\{Z_{ij}\}\right),
\end{array}$$

The dimension $q$ of the latent space corresponds to the number of axes in the PCA or, in other words, to the rank of $\boldsymbol\Sigma = \mathbf{B}\mathbf{B}^\intercal$.


The unkwown parameters are  $\boldsymbol\Theta$ and $\mathbf{B}$, the matrix of .important[_rescaled loadings_]

### Features

  - **Optimization**: _"Similar"_ variational framework (with different gradients)
  - **Model selection**: variational BIC/ICL
      - $\tilde{\text{BIC}}_q = J(\theta, q) - \frac12 \log(n) \left(p (d + q) - q(q-1)/2\right)$
      - $\tilde{\text{ICL}}_q = \tilde{\text{BIC}}_q - \mathcal{H}(q)$
  - **Vizualization:** PCA on the expected latent position $\mathbb{E}_{q}(\mathbf{Z}_i) -\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta  = \mathbf{M}\hat{\mathbf{B}}^\top$

---
# A PCA analysis of the oaks data set

Let us fit PLNPCA on our best model up to now (with TSS as offsets):

```{r PLNPCA offset, cache = TRUE, results = FALSE}
PCA_offset <- 
  PLNPCA(Abundance ~ 1 + offset(log(Offset)), data = oaks, 
         ranks = 1:30, control_main = list(cores = 10))
```


The ouput is of class (`PLNPCAfamily`). It a collection of `R` objects:

```{r print PLNPCA offset}
PCA_offset
```

`PLNPCAfamily` has three methods: `plot`, `getModel`, `getBestModel`<sup>1</sup> 

.footnote[[1] Additional help can be found with `?PLNPCAfamily`, `?getBestModel.PLNPCAfamily`, `?plot.PLNPCAfamily`]

---
# PCA analysis: model selection (I)

The plot function gives you hints about the "right" rank/subspace size of your data

```{r plot PLNPCA offset, fig.width = 12, fig.height=6}
plot(PCA_offset)
```

---
# PCA analysis: model selection (II)

To extract a particular model from the collection, use `getBestModel`:

```{r extract PLNPCA offset}
PCA_offset_BIC <- getBestModel(PCA_offset, "BIC")
```


The extracted object has class `PLNPCAfit`. It inherits from the `PLNfit` class but with additional methods due to its `PCA` nature: when printing `PCA_offset_BIC`, we get

```{r print PLNPCAfit, echo = FALSE}
PCA_offset_BIC
```

---
# PCA analysis: model exploration

Inheritance allows you to rely on the same methods as with `PLN`:

.pull-left[
```{r PLN covariance M01}
corrplot(
  cov2cor(sigma(M01_oaks)),
  tl.cex = .5)
```
]

.pull-right[
```{r PLNPCA covariance}
corrplot(
  cov2cor(sigma(PCA_offset_BIC)),
  tl.cex = .5)
```
]

---
# PCA: vizualisation 

<small>
```{r PCA offset vizu tree, fig.width=6, fig.height=6, fig.align="center"}
factoextra::fviz_pca_biplot(
  PCA_offset_BIC, select.var = list(contrib = 10), addEllipses = TRUE, habillage = oaks$tree,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "tree status") + scale_color_viridis_d()
```
</small>

---
# PCA: removing covariate effects

To hopefully find some hidden effects in the data, we can try to remove confounding ones:

```{r PCA covariate tree, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
PCA_tree <- PLNPCA(Abundance ~ 0 + tree + offset(log(Offset)), 
                   data = oaks, ranks = 1:30, control_main = list(cores = 10))
```

```{r PCA covariate tree plot, echo = FALSE, fig.align="center", fig.width=6, fig.height=6}
PCA_tree %>% getBestModel("BIC") %>% 
factoextra::fviz_pca_biplot(
  select.var = list(contrib = 10), col.ind = oaks$distTOground,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "distance (cm)") + scale_color_viridis_c()
```

---

class: inverse, center, middle

# Discriminant Analysis for counts with PLNLDA

---

# Poisson Discriminant Analysis

PLN-LDA assumes a discrete structure with $K$ groups: the different parameters ${\boldsymbol\mu}_k \in\mathbb{R}^p$ corresponds to the group-specific main effects and the variance matrix $\boldsymbol{\Sigma}$ is shared among groups

$$\begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}(\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta + {\boldsymbol\mu}_k \mathbf{1}_{\{i\in k\}},\boldsymbol\Sigma) \\
  \text{observation space } &  \mathbf{Y}_i | \mathbf{Z}_i \sim \mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right),
\end{array}$$


The unkwown parameters are 
- $\boldsymbol\Theta$, the matrix of regression parameters
- $\mathbf{U} = (\mu_1, \dots, \mu_K)$, the matrix containing the $K$ vectors of group means
- $\boldsymbol{\Sigma}$, the variance-covariance matrix

---
# Geometrical view

```{r PLN_geom_lda_no_offset, echo=FALSE, message=FALSE, fig.width=12, fig.height=8}
set.seed(20171110)
x <- rnorm(100)
y <- rnorm(100)
b <- data.frame(x = x, y = x+y) / 1
group <- sample(LETTERS[1:2], 100, replace = TRUE)
mu <- 0
mean.depth <- rbinom(n = 100, size = 1000, prob = 0.05)
data.perfect <- as.data.frame((b + matrix(rep(mu, each = length(x)), ncol = 2))) %>%
  mutate(group = group,
         x = if_else(group == "A", x+2, x),
         y = if_else(group == "A", y-2, y))
p.latent <- ggplot(data.perfect, aes(x, y, color = group)) + geom_point() + ggtitle(expression(Latent~Space~(Z))) +
  geom_abline(slope = 2, intercept = -3, color = "grey60", linetype = 2) +
  geom_abline(slope = -3, intercept = 1, color = "grey60") +
  # coord_equal() +
  theme(legend.position = "none")
.rpois <- function(lambda) {
  unlist(lapply(lambda, function(x) {rpois(1, x)}))
}
mapped.parameter <- data.perfect %>% mutate_if(is.numeric, exp) # %>%
  # mutate(total.depth = x+y,
  #        x = mean.depth * x / total.depth,
  #        y = mean.depth * y / total.depth)
observation <- mapped.parameter %>% mutate_if(is.numeric, .rpois)
## segment between mapped and observed data
segment.data <- cbind(mapped.parameter %>% select(x, y),
                      observation %>% select(x, y, group))
names(segment.data)[1:4] <- c("x", "y", "xend", "yend")
## Mapped parameters
p.mapped <- ggplot(mapped.parameter, aes(x, y)) + geom_point(aes(col = group)) + ggtitle(expression(Observation~Space~(exp(Z)))) + theme(legend.position = "none")
## Observations only
obs <- group_by(observation, x, y, group)
obs <- dplyr::summarize(obs, count = n())
p.observation.only <- ggplot(obs, aes(x, y, color = group)) +
  geom_point(aes(size = count)) +
  ggtitle(Observation~Space~(Y)~+'noise') +
  theme(legend.position = "none",
        legend.justification = c(1, 1),
        legend.background = element_rect(fill = "transparent"),
        legend.box.background = element_blank())
## Observations and latent parameters
p.observation.mixed <- p.observation.only +
  geom_point(data = mapped.parameter, aes(color = group), alpha = 0.5) +
  geom_segment(data = segment.data, aes(xend = xend, yend = yend), color = "grey60", alpha = 0.2) +
  ggtitle(Observation~Space~+'noise'~(Y==P(exp(Z))))
grid.arrange(p.latent + labs(x = "species 1", y = "species 2"),
             p.mapped  + labs(x = "species 1", y = "species 2"),
             p.observation.mixed + labs(x = "species 1", y = "species 2"),
             p.observation.only + labs(x = "species 1", y = "species 2"),
             ncol = 2)
```

---
# PLNLDA: principles

## Inference

1. Adjust a "standard" PLN with $\mathbf{X} \rightarrow (\mathbf{X}, \mathbf{1}_{\{i\in k\}})$,  $\boldsymbol\Theta \rightarrow (\boldsymbol\Theta, \mathbf{U})$

2. Use estimate $\boldsymbol\Sigma, \mathbf{U}$ and $\boldsymbol\Theta$ and $\tilde{\mathbf{Z}}_i$ to compute
$$\hat{\boldsymbol\Sigma}_{\text{between}} = \frac1{K-1} \sum_k n_k (\hat{\boldsymbol\mu}_k - \hat{\boldsymbol\mu}_\bullet) (\hat{\boldsymbol\mu}_k - \hat{\boldsymbol\mu}_\bullet)^\intercal$$
  - Compute first $K-1$ eigenvectors of $\hat{\boldsymbol\Sigma}^{-1} \hat{\boldsymbol\Sigma}_{\text{between}} = \mathbf{P}^\top \Lambda \mathbf{P}$ (discriminant axes)

## Features

  - **Graphical representation**:  $\tilde{\mathbf{Z}} \mathbf{P} \Lambda^{1/2}$, the coordinates along the discriminant axes
  - **Prediction**: For each group, 
    - Compute (variational) likelihood $p_k = P(\mathbf{Y}_{\text{new}} | \hat{\boldsymbol\Sigma}, \hat{\boldsymbol\Theta}, \hat{\boldsymbol\mu}_k)$
    - Assign $i$ to group with highest posterior probability $\pi_k \propto \frac{n_k p_k}{n}$
  
---

# LDA of the oaks data set

Let us try a PLN-LDA on the `site` variable (`grouping` is a factor of group (or classification) to be considered)

```{r PLNLDA oaks, cache = TRUE, results = FALSE}
myLDA_tree <- 
  PLNLDA(Abundance ~ 1 + offset(log(Offset)), grouping = tree, data = oaks)
```

no need for model selection!

<small>
```{r LDA printing, echo = FALSE}
myLDA_tree
```
</small>

---
# LDA for oaks: covariance model

PLN-LDA can account for various model of the covariance (spherical, diagonal, full)

```{r PLNLDA oaks covariance, results = FALSE, message = FALSE, warning = FALSE, cache = TRUE}
LDA_oaks1 <- PLNLDA(Abundance ~ 1 + offset(log(Offset)), 
  data = oaks, grouping = tree, control = list(covariance = "full"))

LDA_oaks2 <- PLNLDA(Abundance ~ 1 + offset(log(Offset)), 
  data = oaks, grouping = tree, control = list(covariance = "diagonal"))
```

.pull-left[
```{r plot oaks1, echo = FALSE, fig.height = 6}
plot(LDA_oaks1, map = "individual", nb_axes = 1, main = "fully parametrized")
```
]

.pull-right[
```{r plot oaks2, echo = FALSE, fig.height = 6}
plot(LDA_oaks2, map = "individual", nb_axes = 1, main = "diagonal")
```
]

---
# LDA for oaks: prediction

If abundance data for new data are avalaible, their group can be predicted using the `predict` function. We illustrate this on our sample<sup>1</sup> and compare predicted seasons to actual ones

```{r predict PLNLDA season}
predictions <- predict(LDA_oaks1, newdata = oaks, type = "response")
table(predictions, oaks$tree)
```

.footnote[
[1] Predicting the tree of samples used to train the model is bad practice in general and prone to overfitting, we do it for illustrative purposes only. 
]

---

class: center, middle, inverse

# Sparse structure estimation with PLNnetwork <br/> .small[See `r Citet(myBib, "PLNnetwork")`]

---

# Sparse precision for multivariate counts

The PLN-network model add a sparsity constraint on the precision matrix ${\boldsymbol\Sigma}^{-1}\triangleq \boldsymbol\Omega$:

$$\begin{array}{rcl}
  \text{latent space } &   \mathbf{Z}_i \sim \mathcal{N}\left({\mathbf{o}_i + \mathbf{x}_i^\top\boldsymbol\Theta},\boldsymbol\Omega^{-1}\right) &  \|\boldsymbol\Omega\|_1 < c \\
    \text{observation space } &  \mathbf{Y}_i | \mathbf{Z}_i \sim \mathcal{P}\left(\exp\{\mathbf{Z}_i\}\right)
  \end{array}$$


$\rightsquigarrow$ The $\ell_1$-penalty induces selection of direct relations (an underlying network)

## .small[Variational approximation]
  
$$J(\theta, q)  - \lambda  \| \boldsymbol\Omega\|_{1,\text{off}} = \mathbb{E}_{q} [\log p_\theta(\mathbf{Y}, \mathbf{Z})] + \mathcal{H}[q(\mathbf{Z})] - \lambda  \|\boldsymbol\Omega\|_{1, \text{off}}$$

Still bi-concave in $(\boldsymbol\Omega, \boldsymbol\Theta)$ and $(\mathbf{M}, \mathbf{S})$.  

Solving in  $\boldsymbol\Omega$ leads to

$$\hat{\boldsymbol\Omega} = \arg\max_{\boldsymbol\Omega} \frac{n}{2} \left(\log | \boldsymbol\Omega | - \text{trace}(\hat{\boldsymbol\Sigma} \boldsymbol\Omega)\right) - \lambda \|\boldsymbol\Omega\|_{1, \text{off}}: \quad \text{graphical-Lasso problem}$$
with $\hat{\boldsymbol\Sigma} = n^{-1}(\mathbf{M}^\top \mathbf{M} + \mathrm{diag}(\bar{\mathbf{S}}^2)$.

---
# Network inference on the oaks data set

We try the inference of a collections of networks, accounting for the tree effect.

Models are indexed by the sparsity level (or penalty). (E)BIC or similar information criteria can help choosing the right amount.

.pull-left[
```{r PLNnetwork tree, cache = TRUE, results = FALSE}
networks_oaks_tree <- 
  PLNnetwork(
    Abundance ~ 0 + tree + 
           offset(log(Offset)),
    data = oaks
  )
```
]

.pull-right[
```{r plot network family site, fig.width = 7, fig.height = 7, echo = FALSE, caption = "with tree correction"}
plot(networks_oaks_tree)
```
]

---
# PLNnetwork: field access

Let us plot the estimated correlation matrix, after regularization of its inverse, and the corresponding network of partial correlation.

.pull-left[
```{r plot network site, fig.width = 7, fig.height = 7, caption = "with site correction", echo = FALSE}
net2 <- getBestModel(networks_oaks_tree)
corrplot(cov2cor(sigma(net2)), cl.pos = "n",
  is.corr = FALSE, method = 'color')
```
]

.pull-right[
```{r plot net site, fig.width = 8, fig.height = 8, caption = "with site correction", echo = FALSE}
plot(net2)
```
]

---
# PLNnetwork: stability selection

An alternative to model selection criteria is the stability selection  - or StARS in the context of network.

- Basically, it uses resampling to estimate robustness of each edges in the network.
- we keep the value of the penalty that guarantees a given level of robustness in the whole network

.important[Careful!]: its is computationally expensive.

--

In `getBestModel`, when "StARS" is requested, stabiltiy selection is performed if needed:

```{r stability selection, cache = TRUE, eval = FALSE}
net3 <- getBestModel(networks_oaks_tree, "StARS") 
```

---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Allows for covariates
  - Flexible modeling of the covariance structure
  - Efficient VEM algorithm

<br />

## Extensions

- Other covariance structures (spatial, time series, ...)
- Mixture model in the latent space
- Confidence interval and tests for the regular PLN
- Other optimization approaches
     - Exact (composite likelihood, MCMC) starting from the variational solution
     - Stochastic (ADAM, RMSProp, etc: tools from ML optimization) for large scale problems

---

class: center, middle, inverse

# Some comments on data normalization

---
# Abundance data 'normalization'

## Some peculiarities

- Data is noisy and the total number of counts per sample $N_i$ is variable
- $N_i$ is constrained by the technology (ex: DNA sequencer) $\rightarrow$ relative counts

.important[Normalization] aims to correct systematic **uncontrollable** biases such as those induced by sequencing process.
<br />

## How to do so that abundances among different samples are comparable ?

- Transformation with possible imputation of zeros
- Rarefaction or normalization prior to the analysis
- Inclusion of an offset (covariable with a fixed coefficient of $1$) in the model

---
# How to calculate this offset ? (I)

## First idea

`Total Sum Scaling` : uses the total number of counts per sample $N_j$


## Methods from RNA-Seq differential analysis

Assumption: a majority of transcripts is not differentially expressed

Aim : minimizing effect of (very) majority sequences

- Relative Log Expression (`RLE`, Anders and Huber 2010, _DESeq2_)
- Trimmed Mean of M-values (Robinson and Oshlack 2010, _edgeR_)

---
# How to calculate this offset ? (II)

## Methods specifically developped for metagenomics data
Adressing the 'zero-inflation' due to the physical absence or under-sampling of the microbes

- Geometric Mean of Pairwise Ratio method (`GMPR`, Chen et al. 2018)


- Cumulative Sum Scaling (`CSS`, Paulson et al. 2013) (_metagenomeSeq_)

To reduce the influence of OTU highly sampled due to technological sequencing biases, i.e. by being over-represented in the TSS in a sparse environment.

`CSS` selects a scaling factor that is a fixed quantile of OTUs counts.

## User-supplied offsets

---
# Some details

$c_{ki}$: the count of the $k$th OTU $(k = 1, \ldots, q)$ in the $i$th $(i = 1, \ldots, n)$ sample

## Relative Log Expression
Calculates the size factor $s_i$, which estimates the (relative) library size of a given sample, based on

- Step 1: Compute a pseudo-reference sample as the geometric means for all OTUs across samples (less sensitive to extreme value than standard mean)
$\mu_{k}^{GM} = (c_{k1}c_{k2}\ldots c_{kn})^{1/n}, k=1,\ldots,q$

- Step 2: For a given sample,
$s_i= \text{median}_k\{c_{ki}/\mu_k^{GM}\},i=1,\ldots,n$


## Geometric Mean of Pairwise Ratio method

- Step 1: Pairwise comparisons
 $r_{ij} = \underset{k \in \{1, \ldots, p\}|c_{ki}.c_{kj} \ne 0}{\text{Median}} \{ \frac{c_{ki}}{c_{kj}} \}$
- Step 2: Combine pairwise results
$s_i= \left(\prod_{j=1}^n r_{ij}\right)^{1/n},i=1,\ldots,n$

---
# In practice

## _PLNmodels_

```{r offset, eval = FALSE}
prepare_data(counts, covariates, offset = "TSS", ...)
compute_offset(counts, offset = c("TSS", "GMPR", "RLE", "CSS", "none"), ...)
# Specify the use of offset in the formula
PCA_offset <- PLNPCA(Abundance ~ 1 + offset(log(Offset)), oaks, ranks = 1:8)
```

A normalization is .important[necessary] and has often a great impact on downstream analysis.

## How to choose ?

There is .important[no magic recipe] : the 'correct' normalization method to use depends on which assumptions are valid for the biological experiment.

- same / different amount of mRNA / cell
- majority of genes / OTUs is invariant between conditions 
- absence of high count genes / OTUs, similar sampling effort
- existence of controls

---
# References

```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```

