---
title: "PLNmodels"
subtitle: "A collection of Poisson lognormal models <br/> for multivariate analysis of count data"
author: "J. Chiquet, M. Mariadassou, S. Robin<br /><br /> <small>INRAE - Applied Mathematics and Informatics Division</small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://pln-team.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse) # data manipulation
library(corrplot)  # plot of covariance/correlation matrices
```

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)
```

# Reproducibility

## Package PLNmodels

Last stable release of **PLNmodels** on CRAN,  development version available on GitHub.

```{r install, eval=FALSE, tidy=FALSE}
install.packages("PLNmodels")
devtools::install_github("jchiquet/PLNmodels")
```

```{r loading package}
library(PLNmodels)
packageVersion("PLNmodels")
```

## Help and documentation

The [PLNmodels website](https://pln-team.github.io/PLNmodels/) contains

- the standard package documentation 
- a set of comprehensive vignettes for the top-level functions

all formatted with [**pkgdown**](https://pkgdown.r-lib.org)

---

class: inverse, center, middle

# 'Oaks powdery mildew' <br/> A motivating companion data set <br/> .small[See `r Citet(myBib, "JAL16")`]

```{r PLN_motivations, child='motivations.Rmd'}
```

<!-- PLN MODEL -->

---
class: inverse, center, middle

# Multivariate Poisson lognormal models <br/> .small[Statistical framework and inference]

```{r PLN_stat_model, child='PLN_stat_model.Rmd'}
```

---

class: inverse, center, middle

# Multivariate Poisson Regression <br /> .small[Illustrating the oaks powdery mildew data set]

```{r PLN_oaks, child='PLN_oaks.Rmd'}
```

---

# .small[Natural extensions towards multivariate analysis]

  - .important[Classification]: maximize separation between groups with means 

  $$\mathbf{Z}_i \sim \mathcal{N}({\boldsymbol\mu}_k \mathbf{1}_{\{i\in k\}}, \boldsymbol\Sigma), \quad \text{for known memberships}.$$
  - .important[Clustering]: mixture model in the latent space 

  $$\mathbf{Z}_i \mid i \in k  \sim \mathcal{N}(\boldsymbol\mu_k, \boldsymbol\Sigma_k), \quad \text{for unknown memberships}.$$
  
  - .important[Dimension Reduction]: rank constraint matrix $\boldsymbol\Sigma$.
    
    $$\mathbf{Z}_i \sim \mathcal{N}(\boldsymbol\mu, \boldsymbol\Sigma = \mathbf{B}\mathbf{B}^\top), \quad \mathbf{B} \in \mathcal{M}_{pk} \text{ with orthogonal columns}.$$
    
  - .important[Network inference]: sparsity constraint on inverse covariance.
  
  $$\mathbf{Z}_i \sim \mathcal{N}(\boldsymbol\mu, \boldsymbol\Sigma = \boldsymbol\Omega^{-1}), \quad \|\boldsymbol\Omega \|_1 < c.$$
  
  $\rightsquigarrow$ A variant of the variational algorithm is required for each model


---

class: inverse, center, middle

# Classification for counts <br /> PLN discriminant analysis

```{r PLN_LDA_oaks, child='PLN_LDA_oaks.Rmd'}
```

---

class: inverse, center, middle

# Clustering for counts <br /> .small[PLN mixture model]

```{r PLN_mixture_oaks, child='PLN_mixture_oaks.Rmd'}
```

---

class: inverse, center, middle

# Dimension reduction and vizualisation for counts <br/> .small[See `r Citet(myBib, "PLNPCA")`]

```{r PLN_PCA_oaks, child='PLN_PCA_oaks.Rmd'}
```

---

class: center, middle, inverse

# Sparse structure estimation for counts  <br/> .small[See `r Citet(myBib, "PLNnetwork")`]

```{r PLN_network_oaks, child='PLN_network_oaks.Rmd'}
```


---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Flexible modeling of the covariance structure, allows for covariates
  - Efficient V-EM algorithm

## Extensions

- Other variants
      - zero inflation (data with _a lot_ of zeros)
      - covariance structures (spatial, time series, ...)
      - Variable selection ($\ell_1$-penalty on the regression coefficients)
- Other optimization approaches
     - large scale problem: ELBO + SG variant/GPU
     - Composite likelihood, starting from the variational solution
     - .important[log-likelihood] (MCMC, Stochastic gradient)
- Pave the way for Confidence interval and tests for regular PLN


---
# References

```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```

