---
title: "Poisson lognormal models for count data"
subtitle: "Variational inference, Optimization"
author: "J. Chiquet, M. Mariadassou, S. Robin, B. Batardière, J. Kwon, J. Stoehr <br /><br /> <small>MIA Paris-Saclay, AgroParisTech, INRAE</small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://pln-team.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

class: inverse, center, middle

# Multivariate Poisson lognormal models <br/> .small[Motivations, Framework]

```{r setup, echo=FALSE, cache=FALSE, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse)  # data manipulation
library(RefManageR) # reference manager
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)
```

<!-- STATISTICAL MODEL -->
```{r PLN_stat_model, child='PLN_stat_model.Rmd'}
```

---
class: inverse, center, middle

# Variational inference for standard PLN<br/> .small[Optimisation]

<!-- VARIATIONAL INFERENCE -->
```{r PLN_var_inference, child='PLN_var_inference.Rmd'}
```

---
class: inverse, center, middle

# Variational estimators of standard PLN<br/> .small[Properties]

<!-- PROPERTIES OF THE VARIATIONAL ESTIMATORS-->
```{r PLN_var_properties, child='PLN_var_properties.Rmd'}
```

---

# Recent advances <br/> .small[likezlihood maximization, Zero Inflation]

<!-- OPTIMISATION -->
```{r PLN_optim_IS, child='PLN_optim_IS.Rmd'}
```

<!-- ZI-PLN MODEL -->
```{r PLN_ZI, child='PLN_ZI.Rmd'}
```


<!--   - No guarantee for the variational estimators -->
<!--   - No test on the inference parameters -->
<!--   - Environmental Genomics, Single-Cell: more zeros, more rows, more columns -->

<!-- ## Ongoing work -->

<!--   - Zero-inflated version -->
<!--   - Hypothesis testing -->

---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Flexible modeling of the covariance structure, allows for covariates
  - Efficient V-EM algorithm

## Extensions

- Other variants
      - covariance structures (spatial, time series, genetics...)
- Other models
  - Bernoulli/multinomial counterpart to PLN
  - functional data
  - multiple-data integration (e.g., Bernoulli + Poisson)


---
# References

```{r, results='hide', echo = FALSE}
RefManageR::Cite(myBib, 'PLNmodels')
RefManageR::Cite(myBib, 'PLNPCA')
RefManageR::Cite(myBib, 'PLNnetwork')
RefManageR::Cite(myBib, 'Facon2021')
RefManageR::Cite(myBib, 'Lejal2021')
```

<small>
```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```
</small>

