
---

# Oaks powdery mildew data set overview

`r Citet(myBib, "JAL16")` Study .important[effects of the pathogen] _E.Aphiltoïdes_ (mildew) wrt bacterial and microbial communities

## Species Abundances

- .important[Microbial communities] sampled on the surface of $n = 116$ oak leaves
- Communities sequenced and cleaned resulting in $p=114$ OTUs (66 bacteria, 48 fungi).

## Covariates and offsets

Characterize the samples and the sampling, most important being

- `tree`: Tree status with respect to the pathogen (susceptible, intermediate or resistant)
- `distTOground`: Distance of the sampled leaf to the base of the ground
- `orientation`: Orientation of the branch (South-West SW or North-East NE)
- `readsTOTfun`: Total number of ITS1 reads for that leaf
- `readsTOTbac`: Total number of 16S reads for that leaf

---

# Abundance table (I)

```{r glimpse Abundance}
data(oaks)
oaks$Abundance %>% as_tibble() %>% 
  dplyr::select(1:10) %>% 
  rmarkdown::paged_table()
```

---

# Abundance table (II)

```{r glance Abundances, fig.height=6}
log(1 + oaks$Abundance) %>% 
  corrplot::corrplot(is.corr = FALSE,
    addgrid.col = NA,  tl.cex = .5,  cl.pos = "n")
```

---
#  PLN with offsets and covariates (1)

## Offset: .small[modeling sampling effort]

The predefined offset uses the total sum of reads, accounting for technologies specific to fungi and bacteria:

```{r simple PLN offsets, cache = TRUE, results = FALSE}
M01_oaks <- PLN(Abundance ~ 1 + offset(log(Offset)) , oaks)
```

## Covariates: .small[tree and orientation effects ('ANOVA'-like) ]

The `tree` status is a natural candidate for explaining a part of the variance.

- We chose to describe the tree effect in the regression coefficient (mean)
- A possibly spurious effect regarding the interactions  between species (covariance).

```{r PLN covariate oaks, cache = TRUE, results = FALSE}
M11_oaks <- PLN(Abundance ~ 0 + tree + offset(log(Offset)), oaks)
```

What about adding more covariates in the model, e.g. the orientation?

```{r PLN regroup oaks modalities, cache = TRUE, results = FALSE}
M21_oaks <- PLN(Abundance ~  0 + tree + orientation + offset(log(Offset)), oaks)
```

---
#  PLN with offsets and covariates (2)

There is a clear gain in introducing the tree covariate in the model:

```{r PLN covariate oaks results}
rbind(M01 = M01_oaks$criteria,
      M11 = M11_oaks$criteria, M21 = M21_oaks$criteria) %>% 
  knitr::kable(format = "html")
```

Looking at the coefficients $\mathbf{\Theta}$ associated with `tree` bring additional insights:

```{r oaks matrix plot, fig.width=14, fig.height=2, echo = FALSE}
coef(M11_oaks) %>% t() %>% corrplot(method = "color", is.corr = FALSE, tl.cex = 1, cl.pos = "n")
```

---

# Discriminant Analysis

Use the `tree` variable for grouping (`grouping` is a factor of group to be considered)

```{r PLNLDA oaks, results = FALSE}
myLDA_tree <- 
  PLNLDA(Abundance ~ 1 + offset(log(Offset)), grouping = oaks$tree, data = oaks)
```

.pull-left[
```{r plot oaks1, echo = FALSE, fig.height = 6}
plot(myLDA_tree, map = "individual")
```
]

.pull-right[
```{r plot oaks2, echo = FALSE, fig.height = 6}
plot(myLDA_tree, map = "variable")
```
]

---
# A PCA analysis of the oaks data set

```{r PLNPCA offset, cache = TRUE, results = FALSE}
PCA_offset <- 
  PLNPCA(Abundance ~ 1 + offset(log(Offset)), data = oaks, ranks = 1:30)
PCA_offset_BIC <- getBestModel(PCA_offset, "BIC")
```

<small>
```{r PCA offset vizu tree, fig.width=6, fig.height=6, fig.align="center", echo=FALSE}
factoextra::fviz_pca_biplot(
  PCA_offset_BIC, select.var = list(contrib = 10), addEllipses = TRUE, habillage = oaks$tree,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "tree status") + scale_color_viridis_d()
```
</small>

---
# PCA: removing covariate effects

To hopefully find some hidden effects in the data, we can try to remove confounding ones:

```{r PCA covariate tree, cache = TRUE, results = FALSE, warning=FALSE, message=FALSE}
PCA_tree <- 
  PLNPCA(Abundance ~ 0 + tree + offset(log(Offset)), data = oaks, ranks = 1:30)
```

```{r PCA covariate tree plot, echo = FALSE, fig.align="center", fig.width=6, fig.height=6}
PCA_tree %>% getBestModel("BIC") %>% 
factoextra::fviz_pca_biplot(
  select.var = list(contrib = 10), col.ind = oaks$distTOground,
  title = "Biplot (10 most contributing species)"
  ) + labs(col = "distance (cm)") + scale_color_viridis_c()
```

---
# Clustering of the oaks samples

```{r PLNmixture-oaks, cache = TRUE, results='hide'}
PLN_mixtures <-
   PLNmixture(Abundance ~ 1 + offset(log(Offset)), data = oaks, clusters = 1:3)
myPLN_mix <- getModel(PLN_mixtures, 3)
```

.pull-left[
```{r PLN clustering 1 fake, eval = FALSE}
myPLN_mix$plot_clustering_pca()
```

```{r PLN clustering 1, echo = FALSE}
myPLN_mix$plot_clustering_pca(main = 'clustering memberships in individual factor map')
```
]

.pull-right[
```{r PLN clustering 2}
myPLN_mix$plot_clustering_data()
```
]

---
# Network inference

```{r PLNnetwork, cache = TRUE, results='hide', message = FALSE}
networks <- PLNnetwork(Abundance ~ 0 + tree + offset(log(Offset)), data = oaks)
```

.pull-left[
```{r PLNnetwork criteria, echo = FALSE, fig.align='center', fig.height=6}
plot(networks)
```
]

.pull-right[
```{r PLNnetwork network, echo = FALSE, fig.align='center', fig.height=6}
type <- rep(c("bacterial", "fungal", "Alphiltoides"), c(66,114-67,1))
net <- fancy_network(getBestModel(networks, "EBIC"), type)
print(net)
```
]


