---
title: "PLNmodels"
subtitle: "A collection of Poisson lognormal models <br/> for multivariate analysis of count data"
author: "J. Chiquet, M. Mariadassou, S. Robin, B. Batardière, J. Kwon<br /><br /> <small>Paris-Saclay, AgroParisTech, INRAE</small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://pln-team.github.io/PLNmodels"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["pln.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library(tidyverse) # data manipulation
library(corrplot)  # plot of covariance/correlation matrices
library(ggraph)
library(igraph)
library(RColorBrewer)

fancy_network <- function(net, taxonomy, edge_color = c("#F8766D", "#00BFC4")) {
  
  mygraph <- plot(net, plot = FALSE)

  V(mygraph)$type <- as.character(taxonomy)
  E(mygraph)$color   <- factor(ifelse(E(mygraph)$weight > 0, edge_color[1], edge_color[2]), 
                               levels = edge_color)
  
  angle <- 90 - 360 * 1:igraph::gorder(mygraph) / igraph::gorder(mygraph)
  V(mygraph)$hjust <- ifelse(angle < -90, 1, 0)
  V(mygraph)$angle <- ifelse(angle < -90, angle + 180, angle)
  
  p <- ggraph(mygraph, layout = 'linear', circular = TRUE) + 
    geom_edge_arc(aes(edge_colour = color, edge_width = abs(weight)), alpha = 0.5, show.legend = FALSE) +
    geom_node_text(aes(label = type, x = x*1.1, y=y*1.1, angle = angle, hjust = hjust), size = 2, alpha = 1) +
    geom_node_point(aes(colour = type, size = size, alpha = 0.2)) +
    scale_colour_manual(values = c(brewer.pal(6,"Paired"),"maroon4","springgreen", "tan4","plum","brown4","burlywood",brewer.pal(8, "Dark2"), "navy")) +
    #    scale_colour_manual(values = c(brewer.pal(12,"Paired"),brewer.pal(8, "Dark2"), "navy")) +
    scale_size_continuous( range = c(0.1,10) ) +
    theme_void() +
    theme(plot.margin=unit(c(0,0,0,0),"cm"),) + 
    guides(alpha = "none", size = "none") + 
    expand_limits(x = c(-1.5, 1.5), y = c(-1.5, 1.5))
  
  invisible(p)
}
```

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./PLNreferences.bib", check = FALSE)
```

# Reproducibility

## R/C++ Package PLNmodels

Last stable release on CRAN,  development version available on GitHub.

```{r install, eval=FALSE, tidy=FALSE}
install.packages("PLNmodels")
remotes::install_github("PLN-team/PLNmodels@dev")
```

```{r loading package, message=FALSE}
library(PLNmodels)
packageVersion("PLNmodels")
```

```{r future, echo = FALSE}
options(future.fork.enable = TRUE)
future::plan("multicore", workers = 10)
```

## Python module

A Python + PyTorch implementation is coming

## Advertisement .small[(more, sorry about that)]

[https://computo.sfds.asso.fr](https://computo.sfds.asso.fr), a new journal promoting reproducible research

---
# Resources

## Help and documentation

The [PLNmodels website](https://pln-team.github.io/PLNmodels/) contains the standard package documentation and a set of comprehensive vignettes for the top-level functions

## Publications

```{r, results='hide', echo = FALSE}
RefManageR::Cite(myBib, 'PLNmodels')
RefManageR::Cite(myBib, 'PLNPCA')
RefManageR::Cite(myBib, 'PLNnetwork')
RefManageR::Cite(myBib, 'Facon2021')
RefManageR::Cite(myBib, 'Lejal2021')
```

<small>
```{r, 'PLNrefs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```
</small>

---

class: inverse, middle

# PLNmodels: what is done<sup>1</sup> 

  1. .large[Motivations]
  2. .large[A family of models for multivariate analysis]
  3. .large[Efficient variational inference]
  4. .large[Illustration]

.footnote[[1] and published]

<!-- STATISTICAL MODEL -->
```{r PLN_stat_model, child='PLN_stat_model.Rmd'}
```

<!-- VARIATIONAL INFERENCE -->
```{r PLN_var_inference, child='PLN_var_inference.Rmd'}
```

<!-- ILLUSTRATION -->
```{r PLN_oaks_analysis, child='PLN_oaks_analysis.Rmd'}
```

---

class: inverse, middle

# Some more recent research

## Limitations

  - No guarantee for the variational estimators
  - No test on the inference parameters
  - Environmental Genomics, Single-Cell: more zeros, more rows, more columns
  
## Ongoing work

  - Zero-inflated version
  - Hypothesis testing
  - Large scale and/or exact optimization
  
<!-- ZI-PLN MODEL -->
```{r PLN_ZI, child='PLN_ZI.Rmd'}
```

<!-- TESTS -->
```{r PLN_test, child='PLN_tests.Rmd'}
```

<!-- NEW OPTIM -->
```{r PLN_optim, child='PLN_optim.Rmd'}
```


```{r future off, echo=FALSE}
future::plan("sequential")
```

---
# Conclusion

## Summary

  - PLN = generic model for multivariate count data analysis
  - Flexible modeling of the covariance structure, allows for covariates
  - Efficient V-EM algorithm

## Extensions

- Other variants
      - covariance structures (spatial, time series, genetics...)
- Other models
  - Bernoulli/multinomial counterpart to PLN
  - functional data
  - multiple-data integration (e.g., Bernoulli + Poisson)


---
# References

```{r, 'refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```

